A Standard for the Scholarly Citation of Archaeological Data

### 2017-09-07 analysis/paper/paper.Rmd
add some studies of PLOS One sharing

While public and community benefits are well-rehearsed and widely agreed on (Table X), they offer no, or only weak, tangible incentives to the individual researcher. Previous work has identified citation advantages as a tangible benefit to  individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table X for a summary). Increased publication productivity also appeared to be associated with data sharing behaviours. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only 5 for projects without archived data.

@@

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and recieved `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We recieved `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We interpret these to mean that the author does not have time to prepare their data for sharing, perhaps because the data are not already in a well-organised for ready for distribution. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing. Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the one PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014): "Publication is conditional upon the agreement of the authors to make freely available any materials and information described in their publication that may be reasonably requested by others for the purpose of academic, non-commercial research."

We see similar rates of sharing in response to email requests that have been described in other disciplines. However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses suggests two key reasons why author are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data, and that their data is not in a form suitable for sharing and the task of organising those data is not a high priority. These reasons not to share have also been documented in other research areas. The absence of patterns in the date of the paper, student status of the author and other variables may relate to more to our small sample size than a real absence of relationships in data sharing behaviour.

@@

Our third investigation of archaeologists' data sharing behaviours is an analysis of all records (n = `r prettyNum(nrow(dc_arch), big.mark = ",")`, as of August 2017) of archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The majority of records come from the Archaeology Data Service (ADS) in the UK, followed by the Dutch Archiving and Network Services (DANS-EASY). Although the earliest data dates to `r first_date`, file deposition started and steadily increased during 1990-2008, as personal computers become more affordable and data easier to digitize. There is a substantial increase in 2009 when the DANS-EASY repository saw in influx of records from Dutch CRM firms. This relates to the introduction of Dutch archaeology legislation, the _Wet op de archeologische monumentenzorg_ (Wamz), in 2007 that formally obligated archaeologists in the Netherlands to archive their data via DANS within two years of completion of their projects (Keers et al. 2011). After 2009 the volume of deposits declined in many repositories, with the exception of generic repositories such as Figshare, Zenodo and ResearchGate, which continued to grow.

@@

Data sharing is a relatively new behaviour for most archaeologists, so it is important to clarify what does not count as data sharing. Making data available on a web site that is not a trustworthy, persistent-identifier-issuing repository is not data sharing. For example, data files hosted on a personal faculty webpage or the web server of a research project is problematic because these files could disappear at any tim, the data can change in untracked ways, and published links to the files are broken when these websites are reorganised, making the files inaccessible. Similarly, authors who write in their publications that 'the data are available by request' are not sharing data. Not only are the data not actually available, but there is no way to enforce that the authors will actually honour any requests for their data. When archaeological data is available to share, the first considration huold be deposit in a repository....

@@

@@

@@

@@

@@

>>>>

While public and community benefits are well-rehearsed and widely agreed on (Table X), they offer no, or only weak, tangible incentives to the individual researcher (Rowhani-Farid et al. 2017). Previous work has identified citation advantages as a tangible benefit to individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table X for a summary). Increased publication productivity also appeared to be associated with data sharing behaviours. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only 5 for projects without archived data.

@@

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and recieved `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We recieved `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We interpret these to mean that the author does not have time to prepare their data for sharing, perhaps because the data are not already in a well-organised for ready for distribution. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing.

Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the one PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014): "Publication is conditional upon the agreement of the authors to make freely available any materials and information described in their publication that may be reasonably requested by others for the purpose of academic, non-commercial research." Chambers' (2017) examined 50 PLOS One papers on brain images published in November 2014 - May 2015 and found that only 38% of these papers archived their data in an open repository. Similarly, Nuijten et al. (2017) found that of 462 PLOS One articles that promised data availability, 29% of these did not have any available data. Unfulfilled promised to share data are not unique to PLOS One (Kidwell et al. 2016), but these studies highlight the need to resolve ambiguity among authors about what counts as data, and for editors and peer reviwers to more critically assess whether an author has been made their data openly available.

We see similar rates of sharing in response to email requests among archaeologist to what have been described in other disciplines (Tenopir et al. 2015; Vanpaemel et al. 2015). However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses suggests two key reasons why author are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data, and that their data is not in a form suitable for sharing and the task of organising those data is not a high priority. These reasons not to share have also been documented in other research areas. The absence of patterns in the date of the paper, student status of the author and other variables may relate to more to our small sample size than a real absence of relationships in data sharing behaviour.

@@

Our third investigation of archaeologists' data sharing behaviours is an analysis of all records (n = `r prettyNum(nrow(dc_arch), big.mark = ",")`, as of August 2017) of archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The majority of records come from the Archaeology Data Service (ADS) in the UK, followed by the Dutch Archiving and Network Services (DANS-EASY). Although the earliest data dates to `r first_date`, file deposition started and steadily increased during 1990-2008, as personal computers become more affordable and data easier to digitize. There is a substantial increase in 2009 when the DANS-EASY repository saw in influx of records from Dutch CRM firms. This relates to the introduction of Dutch archaeology legislation, the _Wet op de archeologische monumentenzorg_ (Wamz), in 2007 that formally obligated archaeologists in the Netherlands to archive their data via DANS within two years of completion of their projects (Keers et al. 2011). Anothe spike occurs in 2012 when tDAR had a large increate in deposits, likley due to them waiving their deposit fees for that year (Ellison 2012). After 2009 the volume of deposits declined in many repositories, with the exception of generic repositories such as Figshare, Zenodo and ResearchGate, which continued to grow.

@@

Data sharing is a relatively new behaviour for most archaeologists, so it is important to clarify what does not count as data sharing. Making data available on a web site that is not a trustworthy, persistent-identifier-issuing repository is not data sharing. For example, data files hosted on a personal faculty webpage or the web server of a research project is problematic because these files could disappear at any tim, the data can change in untracked ways, and published links to the files are broken when these websites are reorganised, making the files inaccessible. Sharing data via supplementary files attached to a journal article is problematic because the are inaccessible to non-subscribers for paywalled journals, and because publishers often alter the data file formats and and file names during the production process. For example, machine-readable format such as CSV are sometimes presented as PDF or Microsoft Word files in supplementary files. The modification of supplementary file names during the article production process can break scrips that attempt to read files by referring to file names as they were prior to the publisher handling the files.  Finally, authors who write in their publications that 'the data are available by request' are not sharing data. Not only are the data not actually available, but there is no way to enforce that the authors will actually honour any requests for their data.  A simple strategy for avoiding these pseudo-sharing behaviours is to deposit data files in a trustworthy data reposity, and link to the files in reports and manuscripts.

@@

Chambers, C. (2017). The seven deadly sins of psychology: A manifesto for reforming the culture of scientific practice. Princeton, NJ: Princeton University Press.

@@

Ellison, L. A. (2012). Just 20 days left!! Upload your archaeological information FREE to tDAR!! | The Digital Archaeological Record. Retrieved September 7, 2017, from https://www.tdar.org/news/2012/12/just-20-days-left-upload-your-archaeological-information-free-to-tdar/

@@

Nuijten, M. B., Borghuis, J., Veldkamp, C. L. S., Alvarez, L. D., van Assen, M. A. L. M., & Wicherts, J. M. (2017, July 13). Journal Data Sharing Policies and Statistical Reporting Inconsistencies in Psychology. Retrieved from psyarxiv.com/sgbta

@@

Rowhani-Farid, A., Allen, M., & Barnett, A. G. (2017). What incentives increase data sharing in health and medical research? A systematic review. Research Integrity and Peer Review, 2(1), 4. https://doi.org/10.1186/s41073-017-0028-9

@@

Tenopir, C., Dalton, E. D., Allard, S., Frame, M., Pjesivac, I., Birch, B., … Dorsett, K. (2015). Changes in Data Sharing and Data Reuse Practices and Perceptions among Scientists Worldwide. PLOS ONE, 10(8), e0134826. https://doi.org/10.1371/journal.pone.0134826
